On the Minimum Core Mass for Giant Planet Formation at Wide Separations

### 2014-02-14 tex/PaperI/ms.tex
updated tex and response, mostly minor edits

This work develops quasi-static models in which the core has a fixed mass and is no longer accreting solids.  By isolating the process of KH contraction, we determine the minimum timescale for runaway atmospheric growth, and thus for giant planet formation by core accretion.  Previous studies have considered this type of growth \citep[hereafter I00; PN05, respectively]{ikoma00, pn05}, which is an end-member case of phases 2 and 3 in more detailed quasi-static models.  Especially since phase 2 atmospheric growth is often the slowest stage of core accretion, understanding the fastest allowed rates of gas accretion is of crucial importance.  Our study develops simplified cooling models for the purposes of broadly exploring parameter space -- especially in the outer disk -- and developing a more detailed understanding of atmospheric accretion.

@@

To facilitate comparison with our numerical results, \textbf{we rescale the analytic results as follows.   Since growth times are too slow in the non-self-gravitating analytic model, we modify the runaway grown criterion to occur at an effective crossover mass $M_{\rm atm} = f M\co$ with $f < 1$.  Specifically we choose $f = 0.13$, because with this value the analytic model gives the same critical core mass at 10 AU as the numerical model, for our standard choices of other parameters.   This prescription does not mean that runaway growth physically occurs at low atmosphere masses, nor does it cause perfect agreement between the models.  Rather by artificially accelerating growth in the analytic model, we can more easily compare the parameter scalings of the two models.  Mathematically, the modified crossover mass amounts to replacing $\xi \rightarrow f\xi$ in our analysis.}  With $f = 0.13$, the revised scalings are
\begin{subeqnarray}
t_{\rm run} & \approx & 3 \times 10^6 {F_T^{5/2}  F_\kappa \left(\xi \over 3.4\right)^2  \over \mcn{10}^{5/3} \aun{10}^{15 / 14}} \yr  \slabel{eq:tcoolanf} \\
\MC & \approx & 8\, {F_T^{3/2} F_\kappa^{3/5}   \left(\xi \over 2.6 \right)^{6/5} \over \aun{10}^{9 / 14}} \; M_\oplus\ \, ,  \slabel{eq:Mcritf}
\end{subeqnarray} 
now using $t_{\rm run}$, defined as the time when $M_{\rm atm} = f M\co$,  instead of $t_{\rm co}$ for the runaway growth timescale.  When plotting these rescaled analytic results, we include the variations in $\xi$ from \Eq{eq:xidef}.

@@

The accelerated growth in the numerical model, as shown in the bottom panel of \Fig{fig:Ltplot}, is also a direct result of the higher cooling luminosities with self-gravity included.  \textbf{Even when $M_{\rm atm} / M\co$ is only few percent, i.e.\ well before the crossover mass, the effect of self-gravity is  quite evident.  Closer to the star, the effects of self-gravity are not as strong for low atmosphere masses.  Nevertheless our models all show that self-gravity noticeably accelerates growth for $M_{\rm atm} > 0.1 M\co$.}

@@

The effect of mean molecular weight is also shown in \Fig{fig:tvsMcomp}.  A lower $\mu$ gives longer growth times, because more cooling is required to compress the atmosphere.  \textbf{This effect is both well established in core accretion studies \citep{stevenson82} and intuitive since the atmospheric scale height $\propto 1/\mu$ is more extended for lower $\mu$.   The Bondi radius decreases as $\RB \propto \mu$ giving a smaller gravitational sphere of influence (and a weaker compression at $\RH$ when that is the more relevant scale).} For $\mu = 2.0$, which represents the idealized case of an H$_2$ atmosphere completely devoid of Helium, $t_{\rm run}$ increases by factors of $\sim 2 - 3$.  Thus fairly drastic changes in atmospheric composition are required for $\mu$ to significantly affect core accretion timescales. \textbf{In principle, changes in the mean molecular weight of the gas also affect the EOS, including $\delad$, but such effects are not considered here, see \S\ref{sec:EOS}.}

@@

We can directly compare our results with other studies of protoplanetary atmospheric growth in the absence of planetesimal accretion, notably those in I00 and PN05.  A major distinguishing feature of our study is that we study a range of disk conditions, while I00 and PN05 focused on growth at the location of Jupiter, 5.2 AU, in their disk models.  Also both I00 and PN05 solved the full time dependent energy equation and considered more detailed EOSs, in addition to other detailed differences in model parameters.  Despite these differences, the qualitative and quantitative agreement with our study is good.

@@

One goal of our study was to explain the extended luminosity minimum in evolutionary models that characterizes phase 2 in core accretion models as described in the introduction.  Previous work (see Figure 2 in I00 and Figures 3 and 4 in PN05) show that this luminosity minimum is an intrinsic feature of atmospheric cooling even in the absence of planetesimal accretion.  We not only reproduce this effect in a simple model (see \Figs{fig:Ltplot}{fig:LtvsMopacity}) but show that self-gravity is the essential ingredient to produce a broad luminosity minimum well before the crossover mass.

@@

\textbf{A direct comparison with studies that only include planetesimal accretion, and neglect atmospheric cooling, is difficult.  Note however that I00 establishes the correspondence between the minimum luminosity in evolutionary models and the minimum planetesimal accretion rate needed in static models, which gives rise to the classical critical core mass of static models \citep{mizuno78, stevenson82}.}

\textbf{In terms of modern static studies, our results complement (and borrow some tools from) R06 and \citet{Raf11}.  These studies considered the disk radius dependence of core accretion for protoplanets that continuously accrete planetesimals.  We show that the limits on core accretion at large radial distance which \cite{Raf11} (and references therein) claim can be overcome if planetesimal accretion shuts off and the atmosphere is allowed to cool.  The plausibility of rapid core growth followed by negligible subsequent solid accretion admittedly remains highly uncertain, as discussed in more detail in the Summary.}

\section{Neglected Effects}\label{sec:neglected}
\textbf{Since one goal of this paper was to obtain a detailed understanding of atmospheric evolution with a simple model, we have necessarily ignored many  effects of potential significance.  We briefly address the most important of these and note that a followup work, Piso, Youdin \& Murray-Clay (in prep., hereafter Paper II), will extend these models to address some of them.}

@@

\textbf{A stronger constraint is that} core growth subsequently slow severely, to allow the atmosphere to cool and contract.  To be more quantitative, the minimum cooling luminosity in \Fig{fig:Ltplot}, $L\approx 3.5 \times 10^{24}\;\mathrm{erg s}^{-1}$, could be cancelled by low levels of heating from solid accretion.   A core mass doubling timescale of $\sim$$400$ Myr, or faster, would thus provide enough heating to stall atmospheric cooling and growth.   \textbf{An additional concern is that isolation masses tend to grow with disk radius, as $M_{\rm iso} \propto \varSigma_{\rm p}^{3/2} a^2 \propto a^{3/4}$ under the approximation that the surface density of accreted planetesimals ($\varSigma_{\rm p}$) scales with the gas \citep{youdin13}.   While this behavior is nominally inconsistent with final core masses that decline with distance, the predictive power of the isolation mass is imperfect.   For starters, the efficiency of planetesimal formation remains uncertain.  Moreover the locality of core growth, which underlies the isolation mass, disappears when accreted solids drift and/or cores migrate significantly. }

>>>>

This work develops quasi-static models in which the core has a fixed mass and is no longer accreting solids.  By isolating the process of KH contraction, we determine the minimum timescale for runaway atmospheric growth, and thus for giant planet formation by core accretion.  Previous studies have considered this type of growth \citep[hereafter I00; PN05, respectively]{ikoma00, pn05}, which is an end-member case of phases 2 and 3 in more detailed quasi-static models.  Especially since phase 2 of atmospheric growth is often the slowest stage of core accretion, understanding the fastest allowed rates of gas accretion is of crucial importance.  Our study develops simplified cooling models for the purposes of broadly exploring parameter space -- especially in the outer disk -- and developing a more detailed understanding of atmospheric accretion.

@@

To facilitate comparison with our numerical results, \textbf{we rescale the analytic results as follows.   Since growth times are too slow in the non-self-gravitating analytic model, we modify the runaway growth criterion to occur at an effective crossover mass $M_{\rm atm} = f M\co$, with $f < 1$.  Specifically, we choose $f = 0.13$, because with this value the analytic model gives the same critical core mass at 10 AU as the numerical model, for our standard choices of other parameters.   This prescription does not mean that runaway growth physically occurs at low atmosphere masses, nor does it cause perfect agreement between the models.  Rather by artificially accelerating growth in the analytic model, we can more easily compare the parameter scalings of the two models.  Mathematically, the modified crossover mass amounts to replacing $\xi \rightarrow f\xi$ in our analysis.}  With $f = 0.13$, the revised scalings are
\begin{subeqnarray}
t_{\rm run} & \approx & 3 \times 10^6 {F_T^{5/2}  F_\kappa \left(\xi \over 3.4\right)^2  \over \mcn{10}^{5/3} \aun{10}^{15 / 14}} \yr  \slabel{eq:tcoolanf} \\
\MC & \approx & 8\, {F_T^{3/2} F_\kappa^{3/5}   \left(\xi \over 2.6 \right)^{6/5} \over \aun{10}^{9 / 14}} \; M_\oplus\ \, ,  \slabel{eq:Mcritf}
\end{subeqnarray} 
now using $t_{\rm run}$, defined as the time when $M_{\rm atm} = f M\co$,  instead of $t_{\rm co}$ for the runaway growth timescale.  When plotting these rescaled analytic results, we include the variations in $\xi$ from \Eq{eq:xidef}.

@@

The accelerated growth in the numerical model, as shown in the bottom panel of \Fig{fig:Ltplot}, is also a direct result of the higher cooling luminosities with self-gravity included.  \textbf{Even when $M_{\rm atm} / M\co$ is only few percent, i.e.\ well before the crossover mass, the effect of self-gravity is  quite evident.  Closer to the star, the effects of self-gravity are not as strong for low atmosphere masses.  Nevertheless, all our models show that self-gravity noticeably accelerates growth for $M_{\rm atm} > 0.1 M\co$.}

@@

The effect of mean molecular weight is also shown in \Fig{fig:tvsMcomp}.  A lower $\mu$ gives longer growth times, because more cooling is required to compress the atmosphere.  \textbf{This effect is both well established in core accretion studies \citep{stevenson82} and intuitive since the atmospheric scale height $\propto 1/\mu$ is more extended for lower $\mu$.   The Bondi radius decreases as $\RB \propto \mu$, giving a smaller gravitational sphere of influence (and a weaker compression at $\RH$ when that is the more relevant scale). The larger atmospheric scale height due to a lower $\mu$ also results in a larger pressure depth at the RCB, $P\cb$, and thus a lower $R\cb$, cf. \Eq{eq:PcbRcb}.} For $\mu = 2.0$, which represents the idealized case of an H$_2$ atmosphere completely devoid of Helium, $t_{\rm run}$ increases by factors of $\sim 2 - 3$.  Thus fairly drastic changes in atmospheric composition are required for $\mu$ to significantly affect core accretion timescales. \textbf{In principle, changes in the mean molecular weight of the gas also affect the EOS, including $\delad$, but such effects are not considered here (see \S\ref{sec:EOS}).}

@@

\textbf{We can directly compare our results with other studies of protoplanetary atmospheric growth in the absence of planetesimal accretion, notably those in I00 and PN05.  A major distinguishing feature of our study is that we explore a range of disk conditions, while I00 and PN05 focus on growth at the location of Jupiter, 5.2 AU, in their disk models.  Moreover, both I00 and PN05 solve the full time dependent energy equation and consider more detailed EOSs, in addition to other detailed differences in model parameters.  Despite these differences, the qualitative and quantitative agreement with our study is good.}

@@

\textbf{One goal of our study was to explain the extended luminosity minimum in evolutionary models that characterizes phase 2 in core accretion models, as described in the introduction.  Previous work (see Figure 2 in I00 and Figures 3 and 4 in PN05) shows that this luminosity minimum is an intrinsic feature of atmospheric cooling even in the absence of planetesimal accretion.  We not only reproduce this effect in our simplified model (see \Figs{fig:Ltplot}{fig:LtvsMopacity}), but we also show that self-gravity is the essential ingredient to produce a broad luminosity minimum well before the crossover mass.}

@@

\textbf{A direct comparison with studies that only include planetesimal accretion, and neglect atmospheric cooling, is difficult.  Note, however, that I00 establishes the correspondence between the minimum luminosity in evolutionary models and the minimum planetesimal accretion rate needed in static models, which gives rise to the classical critical core mass of static models \citep{mizuno78, stevenson82}.}

\textbf{In terms of modern static studies, our results complement (and borrow some tools from) R06 and \citet{Raf11}.  These studies consider the disk radius dependence of core accretion for protoplanets that continuously accrete planetesimals.  We show that the limits on core accretion at large radial distance claimed by \cite{Raf11} (and references therein) can be overcome if planetesimal accretion shuts off and the atmosphere is allowed to cool.  Nevertheless, the plausibility of rapid core growth followed by negligible subsequent solid accretion admittedly remains highly uncertain, as discussed in more detail in \S\ref{sec:conclusions}.}

\section{Neglected Effects}\label{sec:neglected}
\textbf{Since one goal of this paper was to obtain a detailed understanding of atmospheric evolution with a simple model, we have necessarily ignored many  effects of potential significance.  We briefly address the most important of these and note that a followup work, Piso, Youdin \& Murray-Clay (in prep., hereafter Paper II), will extend out current models to address some of these effects.}

@@

\textbf{A stronger constraint is that} core growth subsequently slow severely, to allow the atmosphere to cool and contract.  To be more quantitative, the minimum cooling luminosity in \Fig{fig:Ltplot}, $L\approx 3.5 \times 10^{24}\;\mathrm{erg s}^{-1}$, could be cancelled by low levels of heating from solid accretion.   A core mass doubling timescale of $\sim$$400$ Myr, or faster, would thus provide enough heating to stall atmospheric cooling and growth.   \textbf{An additional concern is that isolation masses tend to grow with disk radius, as $M_{\rm iso} \propto \varSigma_{\rm p}^{3/2} a^2 \propto a^{3/4}$ under the approximation that the surface density of accreted planetesimals ($\varSigma_{\rm p}$) scales with the gas \citep{youdin13}.   While this behavior is nominally inconsistent with final core masses that decline with distance, the predictive power of the isolation mass is imperfect.   For starters, the efficiency of planetesimal formation remains uncertain.  Moreover, the locality of core growth, which underlies the isolation mass, disappears when accreted solids drift and/or cores migrate significantly. }
